---
title: "4th exercise"
author: "Marcell Frisch"
date: '2023 10 25 '
output: html_document

---

# 4 task

## a) part


Sketch the tree corresponding to the partition of the predictor space illustrated on the left of the figure. The
numbers inside the boxes indicate the mean of Y within each region.


```{r}

# Install and load the necessary packages
if (!requireNamespace("DiagrammeR", quietly = TRUE)) {
  install.packages("DiagrammeR")
}

library(DiagrammeR)



```

```{r}

# Create a decision tree structure with circles of the same size and True/False labels
decision_tree <- "
digraph DecisionTree {
  node [shape=circle, style=filled, fillcolor=lightgray, width=1, height=1];
  root [label=\"x1 <= 1\"];
  decision1 [label=\"Mean of Y is 5\"];
  decision2 [label=\"x2 <= 1\"];
  leaf3 [label=\"x1 <= 0\"];
  leaf4 [label=\"Mean of Y is 15\"];
  decision3 [label=\"Mean of Y is 3\"];  
  decision4 [label=\"x2 <= 0\"]; 
  leaf5 [label=\"Mean of Y is 10\"];   
  leaf6 [label=\"Mean of Y is 0\"];   
  root -> decision1 [label=\"False\"];
  root -> decision2 [label=\"True\"];
  
  decision2 -> leaf3 [label=\"True\"];
  decision2 -> leaf4 [label=\"False\"];
  
  leaf3 -> decision3 [label=\"True\"];      
  leaf3 -> decision4 [label=\"False\"];    

  decision4 -> leaf5 [label=\"True\"];   
  decision4 -> leaf6 [label=\"False\"];  
}
"

# Create a graph from the decision tree structure
tree_graph <- grViz(decision_tree)

# Display the decision tree
tree_graph


```

## b) part

Create a diagram similar to the plot on the left in the figure, using the tree illustrated on the right of the
same figure. You should divide up the predictor space into the correct regions, and indicate the mean for
each region. Determine also the fitted function.



```{r}
par(xpd = NA)
plot(NA, NA, type = "n", xlim = c(0, 1), ylim = c(0, 1), xlab = "x1", ylab = "x2", xaxt = "n", yaxt = "n")
lines(x = c(40,40)/100, y = c(0,50)/100)
lines(x = c(0,100)/100, y = c(50,50)/100)
lines(x= c(0,40)/100,y=c(25,25)/100)
lines(x=c(15,15)/100,y=c(25,50)/100)

text(x = 40/100, y = -0.1, labels = c("1"), col = "black") 
text(x = 15/100, y = -0.1, labels = c("0"), col = "black")
text(x = -0.1, y = 25/100, labels = c("1"), col = "black")
text(x = -0.1, y = 50/100, labels = c("2"), col = "black")


text(x = 50/100, y = 75/100, labels = c("Node 9, mean=2.5"),cex=0.5)
text(x = 7.5/100, y = 40/100, labels = c("Node 6, mean=-1.1"),cex=0.5)
text(x = 27.5/100, y = 40/100, labels = c("Node 7, mean=0.12"),cex=0.5)
text(x = 70/100, y = 25/100, labels = c("Node 8, mean=0.4"),cex=0.5)
text(x = 20/100, y = 12.5/100, labels = c("Node 4, mean=-1.8"),cex=0.5)


```




Provides the mean value of y for a specific region.

```{r}
f_function <- function(x1, x2) {
  sapply(1:length(x1), function(i) {
    if (x2[i] < 2) {
      if (x1[i] < 1) {
        if (x2[i] < 1) {
          -1.8
        } else {
          if (x1[i] < 0.003) {
            -1.1
          } else {
            0.12
          }
        }
      } else {
        0.4
      }
    } else {
      2.5
    }
  })
}

result <- f_function(c(-1, 0, 1, 2), c(0, 1, 2, 3))

```


