---
title: "Assignment 2"
author: "Group 2"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

## Task 2

## Task 3

## Task 4

## Task 5

## Task 6

```{r}
library(MASS)

set.seed(123)

# a) draw from standard multivariate normal
X <- mvrnorm(n = 100, mu = rep(0, 100), Sigma = diag(1, nrow = 100, ncol = 100))
names(X) <- paste0("X", 1:100)

# b) use 10 first columns of X and simulated noise to get y
y <- apply(X[, 1:10], MARGIN = 1, FUN = sum) + rnorm(100, mean = 0, sd = sqrt(0.01))

# c) fit LASSO and ridge models with different values of λ 

library(glmnet)

# alpha = 0 is RIDGE, alpha = 1 is LASSO according to elastic net mixing

# glmnet chooses suitable lambda values itself but one could also specify the
# lambda sequence manually (argument: lambda = sequence). We go with the implemented selection of lambda as
# we expect sensible behavior.

RIDGE <- glmnet(x = X, y = y, alpha = 0)

LASSO <- glmnet(x = X, y = y, alpha = 1)
```

```{r}
# d) plot for RIDGE
plot(RIDGE)

# d) plot with xvar = lambda
plot(RIDGE, xvar = "lambda")
```

```{r}
# d) plot for LASSO
plot(LASSO)

# d) plot with xvar = lambda
plot(LASSO, xvar = "lambda")
```
```{r}
# get deviance and lambda
RIDGE_dev_l <- data.frame(deviance = deviance(RIDGE), lambda = RIDGE$lambda)
LASSO_dev_l <- data.frame(deviance = deviance(LASSO), lambda = LASSO$lambda)

# get number of non-zero coefficients and lambda
# step 1 extract coefficients. Rows are Variables, columns correspond to lambda
RIDGE_coef <- coef(RIDGE)
LASSO_coef <- coef(LASSO)

# define an effective 0 as .Machine$double.eps and treat everything smaller as 0
# column sum of logical matrix provides number of non zero coefficients
RIDGE_coef_non_zero <- apply(RIDGE_coef > .Machine$double.eps, MARGIN = 2, sum)
LASSO_coef_non_zero <- apply(LASSO_coef > .Machine$double.eps, MARGIN = 2, sum)

# plot deviance in dependence of lambda
with(RIDGE_dev_l, plot(x = lambda, y = deviance, type = "l"))
with(LASSO_dev_l, plot(x = lambda, y = deviance, type = "l"))

# plot number of non-zeros in dependence of lambda
plot(x = RIDGE$lambda, y = RIDGE_coef_non_zero, type = "l")
plot(x = LASSO$lambda, y = LASSO_coef_non_zero, type = "l")

```


## Task 7


```{r}


```

## Task 8

### a) 

We have to calculate x such that $P(G = 1 \vert x) = P(G = 2\vert x)$. By Bayes' theorem
we get for $G = i$, $i \in \{1, 2\}$

$$
P(G = i \vert x) = \frac{P(x \vert G = i)P(G = 1)}{P(X)}
$$
As $P(G = 1) = P(G = 2)$ we are left with the condition $P(x \vert G = 1) = P(x \vert G = 2)$. But we know that these are just the conditional distributions of $x$ given in the task. Hence we can now equate the log densities of the standard normal and the normal distribution to obtain an expression for the value(s) of $x$ fulfilling the initial equation. We get

$$
  \begin{aligned}
& \frac{1}{\sqrt{2 \pi}} e^{-\frac{1}{2} x^2}=\frac{1}{\sigma \sqrt{2 \pi}} e^{-\frac{1}{2 \sigma^2}(x-\mu)^2}  \\
& \sigma e^{-\frac{1}{2} x^2}=e^{-\frac{1}{2 \sigma^2}(x-\mu)^2} \\
& \log (\sigma)-\frac{1}{2} x^2=-\frac{1}{2 \sigma^2}\left(x^2-2 \mu x+\mu^2\right) \\
& 2 \sigma^2 \log (\sigma)-\sigma^2 x^2=-x^2+2 \mu x-\mu^2 \\
& 0 = (\sigma^2 - 1)x^2 + 2 \mu x - \mu^2 - 2\sigma^2 log(\sigma) \\
\Rightarrow  & x=\frac{-2 \mu \pm \sqrt{4 \mu^2+4\left(\left(\sigma^2-1\right)\left(\mu^2+2 \sigma^2 \log (\sigma)\right)\right.}}{2 x^2+2} \\
& =\frac{-2 \mu \pm 2 \sqrt{\mu^2+\sigma^2 \mu^2-\mu^2+2 \sigma^4 \log (\sigma)-2 \sigma^2 \log (\sigma)}}{2\left(\sigma^2-1\right)} \\
& =\frac{-\mu \pm \sqrt{\sigma^2 \mu^2+2 \log (\sigma)\left(\sigma^4-\sigma^2\right)}}{\sigma^2-i}
\end{aligned}
$$
So we find two points where the probabilities to of $G = 1$ or $G = 2$ are the same for the two normal distrutions. 

### b) 

```{r}
# get boundary points
boundary <- function(mu, sigma) {
  c(x1 = 1/(sigma^2) * (- mu + sqrt(sigma^2 * mu^2 + 2*log(sigma) * (sigma^4 + sigma^2))), 
    x2 = 1/(sigma^2) * (- mu - sqrt(sigma^2 * mu^2 + 2*log(sigma) * (sigma^4 + sigma^2))))
}

b02 <- boundary(mu = 0, sigma = sqrt(2))
b11 <- boundary(mu = 1, sigma = sqrt(1))
```


### c)

```{r}
library(tidyverse)
library(ggthemes)
# draw from standard normal 
n = 10^6
sn <- rnorm(n)

# draw from mu = 0, sigma^2 = 2
n02 <- rnorm(n, mean = 0, sd = sqrt(2))

# draw from mu = 1, sigma^2 = 1
n11 <- rnorm(n, mean = 1, sd = 1)

df <- data.frame(sn = sn, n02 = n02, n11 = n11) %>% 
  pivot_longer(cols = c(n02, n11))

# plot the densities (boundaries have to be inserted but not so easy with one 
# ggplot, maybe make two. Then it is just adding to vertical lines.)
ggplot(data = df) +
  geom_density(aes(x = sn)) +
  geom_density(aes(x = value), color = "red") +
  facet_wrap(~ name) +
  theme_base()

```




## Task 9

## Task 10
