---
title: "Assignment 1"
author: "Group 2"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1

## Task 2

Consider a set of training data $\left(\boldsymbol{x}_1, y_1\right), \ldots, \left(\boldsymbol{x}_N, y_N\right)$ drawn at random from a population and some test data $\left(\tilde{\boldsymbol{x}}_1, \tilde{y}_1\right), \ldots, \left(\tilde{\boldsymbol{x}}_M, \tilde{y}_M\right)$ also drawn at random from the same population as the training data are given.

We want to show that: 
\[
\mathrm{E}\left[R_{t r}(\hat{\beta})\right] \leq \mathrm{E}\left[R_{t r}\left(\mathrm{E} \left(\hat{\beta}\right)\right)\right]=\mathrm{E}\left[R_{t e}\left(\mathrm{E} \left(\hat{\beta}\right)\right)\right] \leq \mathrm{E}\left[R_{t e}(\hat{\beta})\right]
\]
where $R_{tr}(\hat{\beta})=\frac{1}{N} \sum_{i=1}^N\left(y_i-\boldsymbol{x}_i^{\top} \hat{\beta}\right)^2$ and $R_{tr}(\hat{\beta})=\frac{1}{M} \sum_{i=1}^M\left(y_i-\boldsymbol{x}_i^{\top} \hat{\beta}\right)^2$

\textbf{Solution:}


1. \textbf{Proving the left inequality:}

   Starting with the definition of $\hat{\beta}$ obtained from the training data:

   \[
   \hat{\beta} = \arg \min_{\beta'} \frac{1}{N} \sum_{i=1}^N\left(y_i-\boldsymbol{x}_i^{\top} \beta'\right)^2
   \]

   Since $\hat{\beta}$ minimizes the training error, we have:

   \[
   R_{tr}(\hat{\beta}) \leq R_{tr}(\beta)
   \]

   Now, taking the expectation of both sides:

   \[
   \mathrm{E}\left[R_{tr}(\hat{\beta})\right] \leq \mathrm{E}\left[R_{tr}(\beta)\right]
   \]

   The random variable $\hat{\beta}$ depends on the training data, but we can take its expectation, resulting in a fixed, non-random vector $\mathrm{E}(\hat{\beta})$. Substituting this into the above inequality:

   \[
   \mathrm{E}\left[R_{tr}(\hat{\beta})\right] \leq \mathrm{E}\left[R_{tr}(\mathrm{E}(\hat{\beta}))\right]
   \]

   This proves the left inequality.
   

2. \textbf{Proving the middle equality:}

   For any fixed vector $\beta$, the expected training error $\mathrm{E}[R_{tr}(\beta)]$ and the expected test error $\mathrm{E}[R_{te}(\beta)]$ are given by:

   \[
   \mathrm{E}[R_{tr}(\beta)] = \frac{1}{N} \sum_{i=1}^N \mathrm{E}\left[(y_i-\boldsymbol{x}_i^{\top} \beta)^2\right] = \mathrm{E}\left[(Y-\boldsymbol{X}^{\top} \beta)^2\right]
   \]

   \[
   \mathrm{E}[R_{te}(\beta)] = \frac{1}{M} \sum_{i=1}^M \mathrm{E}\left[(\tilde{y}_i-\tilde{\boldsymbol{x}}_i^{\top} \beta)^2\right] = \mathrm{E}\left[(\tilde{Y}-\tilde{\boldsymbol{X}}^{\top} \beta)^2\right]
   \]

   This equality holds because both the training and test data come from the same distribution.

3. \textbf{Proving the right inequality:}

   We treat $\hat{\beta}$ as a random vector independent from the test data due to the independence between the training and test data. For this part, we can forget about the training data and think of $\hat{\beta}$ as a random vector independent of the (test) data.

   Starting with the expected test error:

   \[
   \begin{aligned}
   \mathrm{E}[R_{te}(\hat{\beta})] &= \mathrm{E}\left(\tilde{Y}-\tilde{\boldsymbol{X}}^{\top} \hat{\beta}\right)^2 \\
   &= \mathrm{E}\left(\tilde{Y}^2-2 \tilde{Y} \tilde{\boldsymbol{X}}^{\top} \hat{\beta}+\left( \tilde{\boldsymbol{X}}^{\top}\hat{\beta}\right)^2\right) \\
   &= \tilde{Y}^2-2 \tilde{Y}\tilde{\boldsymbol{X}}^{\top} \mathrm{E}\left(\hat{\beta}\right) +\tilde{\boldsymbol{X}}^{\top}\mathrm{E}\left(\hat{\beta}\hat{\beta}^{\top}\right) \tilde{\boldsymbol{X}} \\
   &= \tilde{Y}^2-2 \tilde{Y}\tilde{\boldsymbol{X}}^{\top} \mathrm{E}\left(\hat{\beta}\right) +\tilde{\boldsymbol{X}}^{\top}\left[\mathrm{E}\left(\hat{\beta}\right)\mathrm{E}\left(\hat{\beta}^{\top}\right)+Cov\left(\hat{\beta}\right)\right] \tilde{\boldsymbol{X}}
   \end{aligned}
   \]

   Since the covariance matrix is positive semi-definite, $X^T \operatorname{Cov}(\beta) X \geq 0$, we have:
   
\[
\begin{aligned}
\mathrm{E}[R_{te}(\hat{\beta})] &\geq \tilde{Y}^2-2 \tilde{Y} \tilde{\boldsymbol{X}}^{\top}\mathrm{E}(\hat{\beta})+\tilde{\boldsymbol{X}}^{\top}\mathrm{E}(\hat{\beta})\mathrm{E}(\hat{\beta}^{\top}) \tilde{\boldsymbol{X}} \\
&\geq \left(\tilde{Y}-\tilde{\boldsymbol{X}}^{\top}\mathrm{E}(\hat{\beta})\right)^2
\end{aligned}
\]

   This implies:

   \[
   \mathrm{E}[R_{te}(\hat{\beta})] \geq \mathrm{E}\left(\tilde{Y}- \tilde{\boldsymbol{X}}^{\top}\mathrm{E}\left(\hat{\beta}\right)\right)^2 = \mathrm{E}\left(R_{te}(\mathrm{E}\left(\hat{\beta}\right))\right)
   \]

   Thus, we've proved the right inequality.

Therefore, we have established that:

\[
\mathrm{E}\left[\frac{1}{N} \sum_{i=1}^N\left(y_i-\boldsymbol{x}_i^{\top} \hat{\beta}\right)^2\right] \leq \mathrm{E}\left[\frac{1}{M} \sum_{i=1}^M\left(y_i-\boldsymbol{x}_i^{\top} \hat{\beta}\right)^2\right]
\]


## Task 3

## Task 4

NOT SURE IF ENOUGH! I also tried to calculate both directly but had then problems comparing them.

OLS estimator: $$\hat{\beta}=(X^TX)^{-1}X^Ty$$
Ridge estimator: $$\hat{\beta}^{ridge}(\lambda)=(X^TX+\lambda I)^{-1}X^Ty$$
For $X$ orthonormal we get:\

OLS estimator: $$X^Ty$$
Ridge estimator: $$\hat{\beta}^{ridge}(\lambda)=((1+\lambda)I)^{-1}X^Ty$$
Now we look at the MSE, which we can get by the formula:
$$\mathbb{E}[(\hat{\beta}-\beta)^T(\hat{\beta}-\beta)]$$
Therefore we get:
$$MSE_{OLS}=\mathbb{E}[(X^Ty-\beta)^T(X^Ty-\beta)]$$
$$MSE_{ridge}=\mathbb{E}[(((1+\lambda)I)^{-1}X^Ty-\beta)^T(((1+\lambda)I)^{-1}X^Ty-\beta)]$$
Now for $\lambda=0$ we get that $MSE_{OLS}=MSE_{ridge}$ and otherwise it is always possible to find a $\lambda$, so that we can scale $X^Ty$ to make the MSE of the ridge estimator smaller, as the $MSE_{OLS}$ gets bigger, if $X^Ty-\beta$ gets bigger.

## Task 5

The $k$-nearest-neighbour representation is defined as:
$$\hat f(x_0) = \frac{1}{k}\sum_{i=1}^N  \mathbf{1}_{x_i \in N_k(x_0)}y_i$$
where $N_k(x_0)$ is the neighborhood of $x_0$ defined by the $k$ closest points $x_i$. Therefore the weights are 
$$\ell_i(x_0; \mathcal X) = \frac{1}{k} \mathbf{1}_{x_i \in N_k(x_0)}$$
and the $k$-nearest-neighbor regression is a member of this class.


For the linear regression the we can predict $f$ by 
$$\hat f(x_0) = x_0^T \beta$$
where $\beta = (X^T X)^{-1} X^T y$.  Therefore we have
$$\hat f(x_0) = \sum_{i=1}^N \left( x_0^T (X^T X)^{-1} X^T \right)_i y_i.$$
Hence the weights are
$$\ell_i(x_0; \mathcal X) = \left( x_0^T (X^T X)^{-1} X^T \right)_i$$
and the linear regression is part of this class.

## Task 6

## Task 7

```{r}
# load data
data("diabetes", package = "lars")

# matrices can apparently be columns in df..
# good to know
y <- rnorm(100)
x1 = rnorm(100)
x2 = rnorm(100)
x <- cbind(x1, x2) 
df <- data.frame(y = y, x = I(x))

# but this complicates things so we break up the 
# structure and make a nice df 
# function to get rid of Asis
unAsIs <- function(X) {
    if("AsIs" %in% class(X)) {
        class(X) <- class(X)[-match("AsIs", class(X))]
    }
    X
}

# extract y and x
y <- diabetes$y
x <- unAsIs(diabetes$x)

# make a new df with all the data
diabetes_df <- as.data.frame(cbind(y, x))

```
As instructed, we now set a seed and sample row indices from the set of integers running from 1 to the number of observations with equal probability. 

```{r}
# set seed
set.seed(12)

# split the data into train and test
# step 1: sample 400 indices
ind <- sample(x = 1:nrow(diabetes_df), size = 400)

# subset the datasets as instructed
train <- diabetes_df[ind, ]
test <- diabetes_df[-ind, ]
```

The reason why random sampling is a good idea is that we are not really familiar with the dataset. Specifically we do not know whether observations were sorted by any of the variables available and if so we do not know at all by which one. Just taking the 400 first observations then would lead the information contained in training and test data to be biased by sorting leading ultimately to sampling bias in our estimations. 

INSERT EXPLANATION ABOUT STANDARDIZED VARIABLES HERE.

To analyse the correlation structures we simply calculate a matrix with correlation of all columns in our dataset. The first column contains the correlations of the variables in $X$ with $y$ and the other columns and rows respectively contain the correlations between the columns in $X$. In general high absolute values of $corr(X_k, y)$ are desirable because this implies high co- or countermovement of the dependent and independent variables. This at least hints at predictive power of $X_k$, where $k$ is the column index. Contrary, low values for $corr(X_k, X_j)$, $k \neq j$ are desirable as high values would introduce all the problems associated with multicollinearity, most prominently however the variance of the estimates will become inflated. This means nothing else than a loss in precision of estimates. Another huge problem is that multicollinearity is associated with "almost rank deficient" $X'X$ what can lead to problems if we run our model on a computer system. 

```{r}
library(kableExtra)
# exploration of correlation
correlation_matrix <- round(cor(train), 2)
# eliminate redundancies and make a nice table for the pdf
correlation_matrix[!lower.tri(correlation_matrix)] <- ""
kable(correlation_matrix, booktabs = T)
```


```{r}
# use training data to fit the full model
# get model formula from column names
f <- as.formula(paste0("y~", paste(colnames(train)[-1], collapse = "+")))

fit_full <- lm(data = train, formula = f)

# get variables significant at alpha = 0.05
summary_full <- summary(fit_full)
coefficients <- summary_full$coefficients
significant <- which(coefficients[, 4] < 0.05)[-1]

# in sample MSE
MSE_full_in <- mean(fit_full$residuals^2)

# out of sample MSE
pred_full <- predict(fit_full, newdata = test)
MSE_full_out <- mean((test$y - pred_full)^2)

message(paste("In sample MSE is", MSE_full_in, sep = ": "))
message(paste("Out of sample MSE is", MSE_full_out, sep = ": "))
```

```{r}
# use the significant variables only 
f2 <- as.formula(paste0("y~", paste(colnames(train)[significant], collapse = "+")))

# estimate smaller model
fit_sig <- lm(data = train, formula = f2)

# in sample MSE
MSE_sig_in <- mean(fit_sig$residuals^2)

# out of sample MSE
pred_sig <- predict(fit_sig, newdata = test)
MSE_sig_out <- mean((test$y - pred_sig)^2)

message(paste("In sample MSE is", MSE_sig_in, sep = ": "))
message(paste("Out of sample MSE is", MSE_sig_out, sep = ": "))

# F-Test
anova(fit_full, fit_sig)
```


## Task 8

The Akaike criterion is in general given by 

$$
  AIC = 2K - ln(L)
$$
where $K$ is the number of regressors and $L$ is the likelihood of the model. Since $2K$ is a penalty term, AIC is to be minimized to find the most appropriate model. 

### Best Subset Selection

Best subset selection is basically just getting all $2^P$ possible combinations of explanatory variables available to us and estimating the corresponding regression models. It would be conventient to use the leaps package for this task but there the AIC-criterion is not implemented but only SIC and BIC. So we have to construct the models ourselves, estimate them and calculate AIC.

First we write a function that calculates all possible combinations of regressors and gives back the associated regression formulas. We will not use the leaps and bounds algorithm as the dataset is not to big. 

```{r}
# write function to get formulas
bs_formulas <- function(x = train, dep = "y", intercept_only = T) {
  # extract variables names
  vars <- names(x)
  exps <- vars[vars != dep]
  
  # get all combinations
 f <- lapply(1:length(exps), function(k) {
    # get combinations for given k
    combinations <- combn(exps, m = k, simplify = F)
    # make it a regression formula
    formulas <- lapply(combinations, function(c) {
      paste(dep, paste(c, collapse = "+"), sep = "~")
    })
    # make it a vector again
    unlist(formulas)
    }) 
  # dissolve list again
  output <- unlist(f)
  
  if(intercept_only == T) output <- c(as.formula("y ~ 1"), output)
  
  return(output)
}
```

Then we estimate the models

```{r}
# get formulas
formulas <- bs_formulas() # look at defaults set above

# estimate all models
fits <- lapply(formulas, function(f) lm(data = train, formula = f))

# get log likelihoods
LL <- unlist(lapply(fits, function(f) logLik(f)[1]))

# get number of regressors (K)
K <- unlist(lapply(fits, function(fit) length(fit$coefficients)))

# get AIC
AIC <- 2 * K - 2 * LL

# get the most appropriate model
best_index <- which.min(AIC)

# display it
formulas[[best_index]]
```
Now that we have identified the best model we can assess its in- and out-of-sample performance using MSE again.

```{r}
# get in sample MSE
MSE_BS_in <- mean(fits[[best_index]]$residuals^2)

# get out of sample MSE
MSE_BS_out <- mean((test$y - predict(fits[[best_index]], newdata = test))^2)
```
Finally we conduct an F-Test of the identified model against the full model.

```{r}
anova(fits[[best_index]], fits[[length(fits)]]) # last model is the full one by construction
```

### Backward Stepwise

Here we are lucky because the MASS package provides us with a function that does stepwise regression based on AIC. 

```{r}
library(MASS)

# conduct backwards stepwise regression
backwards_step <- stepAIC(fit_full, direction = "backward")

# get in sample MSE
MSE_backwards_in <- mean(backwards_step$residuals^2)

# get out of sample MSE

MSE_backwards_out <- mean((test$y - predict(backwards_step, newdata = test))^2)

anova(fit_full, backwards_step)
```


## Task 9

## Task 10

