---
title: "Homework 4"
author: "Jacopo Liera"
date: "2023-10-28"
output: pdf_document
---

## Exercise 1

A binary dependent variable is generated by:

$$
Pr(Y = 1|X) = q + (1-2q) \; \cdot  \; \mathbb{1} \left[ \sum^J_{j = 1} X_j > \frac{J}{2} \right]
$$

where $1[]$ is the indicator function$X \sim U(0, 1)^p$, $0 \leq q \leq \frac{1}{2}$,  $J \leq p$ is some predefined (even) number. Describe the probability surface and give the Bayes error rate.


### Solution

The sum of $n$ independent random variables $X_j \sim U(0,1)$ is Irwin-Hall Distributed, with cdf:

$$
F_X = \frac{1}{n!} \sum^{|x|}_{k = 0} (-1)^k {n \choose k} (x- k)^{n-1}
$$

with support $x \in [0, n]$. Also, $E[X] = \frac{n}{2}$. Given the median being also $\frac{n}{2}$, we know the distribution function to be symmetric, further implying that we have 50% probability to get 1 out of the indicator function, and 50% to get a 0, no matter the $J$.

From Chapter 2 we have the Bayes Error defined as:

$$
Err_{Bayes}= 1- E\left[max_{j \in (0,1) } \; P(Y = j|X) \right]
$$
We also have that:
$$
max_{j \in (0,1) } \; P(Y = j|X) =
\begin{cases}
P(Y = 1|X) \;\; if \;\; P(Y =1 |X) > 0.5\\\\
1- P(Y = 1|X) \;\; if \;\; P(Y =1 |X) \leq 0.5
\end{cases}
$$
Since the $\mathbb{1}[]$ takes values of either 0 or 1 50% of the times and $0 \leq q \leq 0.5$, we have that:

$$
max_{j \in (0,1) } \; P(Y = j|X) =
\begin{cases}
P(Y = 1|X) = 1-q \;\; if \;\; \mathbb{1}[] = 1\\\\
P(Y = 0|X) = 1-q \;\; if \;\; \mathbb{1}[] = 0
\end{cases}
$$
Clearly, we have that $max_{j \in (0,1) } \; P(Y = j|X) = 1-q$ and therefore:

$$
Err_{Bayes}= 1- E\left[max_{j \in (0,1) } \; P(Y = j|X) \right] = q
$$

## Exercise 9 


We assume data with the following data generation process:

$$
x = y + \epsilon
$$
where $y$ is a categorical variable with values 1, 2, 3, which occur with equal probability and $\epsilon \sim N(0, 0.2)$ independent.

- Draw 100 data sets of size 100

Here we simply sample the classes in y, the white noise $\epsilon$ to generate the observations $x$.

```{r}
set.seed(137)
data_gen <- function(n_samples){
  
  i <- 1
  vmat <- list()
  
  repeat{
    
    # Break Condition
    if(i == (n_samples + 1)) break
    
    # Generate 100 random numbers
    y <- sample(c(1, 2, 3), size = 100, replace = T)
    eps <- rnorm(100, mean = 0, sd = sqrt(0.2))
    x <- y + eps
    
    # Put it in a matrix
    tmp <- data.frame(x, y, eps)
    vmat[[i]] <-  tmp
    
    i <- i + 1
  }
  
  return(vmat)
}

vmat <- data_gen(n_samples = 100)
```
  
- Determine the sum of the misclassification rates, Gini indices and deviance criteria weighted with the number of observations in each subgroup for the subgroups obtained when splitting the observations using $x$ with thresholds 1.5, 2, and 2.5 and $y$ as dependent variable in the classification problem.

```{r}
# Impurity measures functions
misc_err <- function(p) return( 1 - max(p) )
gini <- function(p) return( sum(p * (1-p)) )
deviance <- function(p) return( (-1) * sum(p * log(p)) )

# Given thresholds
thresholds <- c(1.5, 2, 2.5)

# Function that calculates the impurity given the impurty measure function, 
# the threshold for the split for one single dataset.

impurity_calculate <- function(df, class = "y", col = "x", thresh, FUN){
  
  # Slice the dataset given the threshold into 2 subgroups
  df$split <- cut(df[, col], c(-Inf, thresh, Inf))
  
  # 0 is the <= thresh , 1 is the >thresh
  levels(df$split) <- c(0, 1)
  
  # Number of observations for weighting later on
  nobs <- table(df$split)
  
  # Derive a matrix to count the observations for each split
  class_mat <- table(df[, c("split", class)])
  calc_mat <- t(apply(class_mat, 1, function(x) x/sum(x)))
  
  # Impurity function act here
  res <- apply(calc_mat, 1, FUN)
  
  # Pay attention to NaN Values
  res[is.nan(res)] <- 0
  
  # Weighted impurity measure for the split.
  res_w <- res*nobs
  
  # Return the overall impurity as the sum for each group
  return(sum(res_w))
}

generate_error_tables <- function(impurity_criterion, thresholds, df_list){
  
  j <- 1
  
  out_mat <- matrix(rep(0, 300), nrow = 100)
  colnames(out_mat) <- thresholds
  
  for(t in thresholds){
    
    for(i in 1:length(df_list)){
      
      tmp <- impurity_calculate(df_list[[i]], class = "y", col = "x", thresh = t, FUN = impurity_criterion)
      out_mat[i, j] <- tmp
      
    }
    
    j <- j + 1
    
  }
  
  return(out_mat)
  
}

# Generate error tables: these contain the sum of the errors for each split for each dataset
deviance_table <- generate_error_tables(deviance, thresholds, vmat)
gini_table <- generate_error_tables(gini, thresholds, vmat)
misc_table <- generate_error_tables(misc_err, thresholds, vmat)

sum_table <- rbind(colSums(deviance_table), colSums(gini_table), colSums(misc_table))
rownames(sum_table) <- c("Deviance", "Gini", "Misclassification")

kableExtra::kable(sum_table)
```

- Calculate the best threshold according to each of the three impurity measures for each of the 100 data sets. Summarize and interpret the results.

```{r}
best <- function(vector) as.integer(vector == min(vector)) 

best_deviance <- apply(t(apply(deviance_table, 1, best)), 2, sum)
best_gini <- apply(t(apply(gini_table, 1, best)), 2, sum)
best_misc <- apply(t(apply(misc_table, 1, best)), 2, sum)

best_matrix <- matrix(c(best_misc, best_gini, best_deviance), nrow = 3, byrow = T)
rownames(best_matrix) <- c("Misclasssification", " Gini", "Deviance")
colnames(best_matrix) <- c("1.5", "2", "2.5")

kableExtra::kable(best_matrix)
```

```{r, echo = FALSE}
par(mfrow = c(1, 3))
barplot(best_gini, main = "Best Gini")
barplot(best_deviance, main = "Best Deviance")
barplot(best_misc, , main = "Best Misclassification")
```
From the results, we see that the for both the Deviance and the Misclassification error, the threshold at 2 is performs significantly better, whether for the Gini Index, this is not true anymore and the extreme split points are better performing. One explanation we can give around this, is the fact that for the Misclassification error, the scaled up coefficients turn out to be larger for the subgroup containing the majority of observations, i.e. the ones subgroups $\{obs_i>1.5\}$, $\{obs_i<2.5\}$. Hence, this is clear from the results. We note that the Deviance behaves in a similar way in comparison with the Misclassification error, selecting 2 as the best threshold. The Gini Index is the one which selects the split at 2 the least amount of times. We interpret this with the fact that by construction having one class generally not appearing in the split, is weighted much more in the calculation rather than having a slightly more impure second node. 
